### 2021-07-23 ch/1-introduction.tex
Clean up

\section{Relations to other fields and clarification of terms}
\subsection{aka: Isn't this thesis actually about XY instead of fairness?}
The thesis is principally written from the perspective of algorithmic fairness,
but it touches on other fields as well, like \emph{domain shift} and \emph{causality}.
Furthermore, the thesis makes use of concepts like \emph{transferable representations}
and \emph{interpretability}.
I summarise the areas here and discuss their relevance to the presented work.

@@

The other closely related topic is \emph{causality}.
The argument that a solution to the dataset bias problem involves causality goes as follows:
As mentioned above, the goal is to ignore the bias in the data
and learn the true underlying structure that is hidden in the data.
Often, the most fundamental structure that we can learn is the \emph{causal} structure of the problem of interest.
For example, if an \acs{ML} model truly understood what makes someone a good employee,
it would not need to rely on surface characteristics like whether the applicants name sounds foreign.
% Learning such causal models is not yet completely in the realm of what today's \ac{ML} systems can achieve \citep{pearl2019seven},
% and will likely require a very large amount of data.
% (The models that perhaps come close to this are methods like unsupervised methods like GPT \citep{radford2018improving,radford2019language,brown2020language} that have been trained on enormous amounts of data.)
However, the work in this thesis is not presented under the banner of causality, for several reasons.
% First, a majority of the experiments are on images, such that the input features are pixel values,
% and thus do not lend themselves well to be analysed with methods like Bayesian
First, in as far as the presented methods solve a causal structure problem, they solve a very limited one;
inputs are mapped to outputs, and no detailed exploration of a potential implicit causal model is performed.
Second, the discovered robust structures are not necessarily \emph{causal} in nature, in a narrow sense of the word.
This is especially true in image data:
For example, while discovering that the presence of a smile and the gender of the person in a photograph are two distinct characteristics
requires a deep understanding of human faces,
it does not require knowing how smiles and gender are related via cause-and-effect in the real world.

>>>>

\section{Relations to other fields and clarification of terms}
% \subsection{aka: Isn't this thesis actually about XY instead of fairness?}
The thesis is principally written from the perspective of algorithmic fairness,
but it touches on other fields as well, like \emph{domain shift} and \emph{causality}.
Furthermore, the thesis makes use of concepts like \emph{transferable representations}
and \emph{interpretability}.
I summarise the areas here and discuss their relevance to the presented work.

@@

The other closely related topic is \emph{causality}.
The argument that a solution to the dataset bias problem involves causality goes as follows:
As mentioned above, the goal is to ignore the bias in the data
and learn the true underlying structure that is hidden in the data.
Often, the most fundamental structure that we can learn is the \emph{causal} structure of the problem of interest.
For example, if an \acs{ML} model truly understood what makes someone a good employee,
it would not need to rely on surface characteristics like whether the applicant's name sounds foreign.
% Learning such causal models is not yet completely in the realm of what today's \ac{ML} systems can achieve \citep{pearl2019seven},
% and will likely require a very large amount of data.
% (The models that perhaps come close to this are methods like unsupervised methods like GPT \citep{radford2018improving,radford2019language,brown2020language} that have been trained on enormous amounts of data.)
However, the work in this thesis is not presented under the banner of causality, for several reasons.
% First, a majority of the experiments are on images, such that the input features are pixel values,
% and thus do not lend themselves well to be analysed with methods like Bayesian
First, in as far as the presented methods solve a causal structure problem, they solve a very limited one;
inputs are mapped to outputs, and no detailed exploration of a potential implicit causal model is performed.
Second, the discovered robust structures are not necessarily \emph{causal} in nature, in a narrow sense of the word.
This is especially true in image data:
For example, while discovering that the presence of a smile and the gender of the person in a photograph are two distinct characteristics
requires a deep understanding of human faces,
it does not require knowing how smiles and gender are related via cause-and-effect in the real world.

### 2021-07-23 ch/7-discussion.tex
Clean up

% ======== is it fair that AI has to be fair, if humans tend to be unfair anyhow? ========
%   - yes, because there is no excuse to make something artificially biased, and also, an AI might affect many more people than a single person
%   - it's like asking: isn't it unfair that our president should have a better character than the average person?
It was mentioned before that an \ac{ML} model should be judged by how useful it is,
how it stacks up to realistic alternatives, and not if it is perfect.
However, one has to be careful not to take this philosophy too far.
Namely, one should avoid giving up and saying,
``why should AI need to be fair if humans tend to be bias anyhow?''
The faults of humans has little to do with the question of what we expect of machines.
% and we should certainly not go out of our way to encode the same biases in our creations.
If one of our machines will affect many of our fellow humans,
we would not want it to be harmful, right?
Should we make the machine artificially biased
so that humans do not need to feel too guilty about their own biases?
That way lies folly.
We should always strive to do the best we can.
% ==============

% ======== giving my social perspective viewpoint ========
As AI models (as opposed to mere \ac{ML} models) become more powerful,
the biases as presented in this thesis will likely become less relevant.
However, we can argue that more abstract forms of these problems will still be present.
Spurious correlations in the training data might become less relevant --
because a powerful AI can do its own physics experiments
and thus will be able to figure out the causal structure of our world on its own --
but what is still relevant is the existence of robust, analysable world models,
which it is possible to reason on and do planning with.
The potential trap here is not that the world model is \emph{wrong} per se,
but that it fails to have a useful structure for other tasks.
Furthermore, learning about the physical world will not in and of itself be enough to
become aligned with \emph{human values},
which is here a catch-all term for
what morality is or what good is and what wrong is.
% the things that we want more of, and what bad things are.
Such questions may appear very different from the issue of label bias
and objectives like demographic parity,
but if we take the problem of classical algorithmic fairness to be
that we want the model to treat our fellow humans right,
then this is just the generalisation of that.
% ==============

>>>>

% ======== is it fair that AI has to be fair, if humans tend to be unfair anyhow? ========
%   - yes, because there is no excuse to make something artificially biased, and also, an AI might affect many more people than a single person
%   - it's like asking: isn't it unfair that our president should have a better character than the average person?
It was mentioned before that an \ac{ML} model should be judged by how useful it is,
how it stacks up to realistic alternatives, and not if it is perfect.
However, one has to be careful not to take this philosophy too far.
Namely, one should avoid giving up and saying,
``why should AI need to be fair if humans tend to be bias anyhow?''
The faults of humans has little to do with the question of what we expect of machines.
% and we should certainly not go out of our way to encode the same biases in our creations.
If one of our machines will affect many of our fellow humans,
we would not want it to be harmful, right?
Should we make the machine artificially biased
so that humans do not need to feel too guilty about their own biases?
I believe this to be folly.
We should always strive to do the best we can,
even if this leads to a world in which machines are less biased than humans.
% ==============

% ======== giving my social perspective viewpoint ========
% As AI models (as opposed to mere \ac{ML} models) become more powerful,
% the biases as presented in this thesis will likely become less relevant.
% However, we can argue that more abstract forms of these problems will still be present.
% Spurious correlations in the training data might become less relevant --
% because a powerful AI can do its own physics experiments
% and thus will be able to figure out the causal structure of our world on its own --
% but what is still relevant is the existence of robust, analysable world models,
% which it is possible to reason on and do planning with.
% The potential trap here is not that the world model is \emph{wrong} per se,
% but that it fails to have a useful structure for other tasks.
% Furthermore, learning about the physical world will not in and of itself be enough to
% become aligned with \emph{human values},
% which is here a catch-all term for
% what morality is or what good is and what wrong is.
% % the things that we want more of, and what bad things are.
% Such questions may appear very different from the issue of label bias
% and objectives like demographic parity,
% but if we take the problem of classical algorithmic fairness to be
% that we want the model to treat our fellow humans right,
% then this is just the generalisation of that.
% ==============
